Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-06-25 manuscript/revision2/response_to_reviews/response_to_reviews.qmd

> ::: {.callout-warning appearance="minimal"}
> \- GWAS: The GWAS model the authors use is not specified (nor was it specified in the 2021 Nature paper referenced). I believe looking at the source code that the model is a linear model with PCs to account for structure but no kinship matrix. It’s not clear how many PCs were used and since the raw GWAS results are not presented it’s impossible to tell if this was sufficient to account for the significant structure in these populations. I’m concerned about this because of the extremely unlikely number of significant markers (19K LD blocks). If each of these was a true positive, the average effect size of each block would be only 0.005% of the genetic variance. With only 350 accessions, the power to detect a locus that explains 10% of the genetic variance is only 50% with alpha = 10\^-5 and MAF = 0.05. So if the model is detecting tons of loci it’s likely that the model is severely biased by population structure, and these biases are likely to be somewhat consistent across gardens because the structure is similar. mash doesn’t help when the individual models are biased, it’ll find patterns of correlation whether they are biological or not and use those to shrink effect sizes together.
> :::

First, we now better document our GWAS methodology for this manuscript in Section S2 of the SI Appendix. We also include additional information on GWAS, including the number of PCs used and the Manhattan and QQ plots and associated data tables in the Github repository associated with the paper.

Second, though we selected 19K LD blocks in our set of 'strong' effects for *mash*, very few of these effects were significant (\<500 relatively unlinked LD blocks per set of eight gardens; this still represents some genomic inflation, no doubt, but not nearly as bad as 19K significant blocks would be). We now clarify this important detail in the main manuscript and in the SI Appendix.

Third, population structure could certainly bias the resultant mash models. We describe the steps we take to prevent "garbage in" to our mash models, including removing conditions where GWAS had significant population structure. However, it is our intuition that, should biases remain in our models, as they undoubtedly do, mash would not load these biased effects onto our GxWeather covariance matrices - instead, we these effects could perhaps load onto canonical matrices, such as garden-specific effects, or perhaps onto data-driven matrices. Thus, we expect our assessment of the proportion of the genome that shows GxWeather effects to be conservative if there is residual population structure.

> ::: {#comment-cov-correlations .callout-warning appearance="minimal"}
> \- Mash: A lot of the results interpretation is based on interpreting the SNP loadings on the specified covariance matrices. This is a secondary use of mash (the primary being the refinement of effect sizes), and while they do interpret these somewhat in the mash paper, these loadings need to be interpreted with caution. If these covariance matrices are similar, mash will somewhat arbitrarily assign weight to any of the matrices because all lead to the same fit to the effect-size data. The mash algorithm doesn’t give the full posterior distribution on the loadings so you can’t check for posterior correlations there. It seems likely to me that the hypothesis-driven and data-driven covariance matrices are somewhat correlated here, and the correlations may differ between green-up and flowering because of the better correlation between the environmental metrics and flowering than green-up. In the mash paper, they used cross-validation with the likelihood in the testing set as the evaluation metric to compare models. It might be safer to try dropping specific covariance matrices and comparing the model performance in a held-out testing set to evaluate the importance of the hypothesis-derived covariance matrices.
> :::

@@

In practice this greedy algorithm approach offers a fast way to identify the point where redundancy in the addition of a new covariance matrix results in no change to the likelihood of the mash model. Importantly, this removes the potential for arbitrary assignment of weights to the covariance matrices in the model and allows for the selection of matrices that most accurately capture the underlying biology captured by the site-specific effect sizes. We applied this algorithm to only the hypothesis matrices for each phenotype (Figures 1 and 2) using the previously used sets of significantly associated and randomly selected variants in the Gulf subpopulation individuals, Midwest subpopulation individuals, and the two subpopulations combined. The results indicate that only a small number of the original hypothesis matrices (between three and five) are necessary to reach the maximum likelihood model when using the same likelihood ratio test implemented in Urbut et al., indicating that correlation among hypothesis covariance matrices is high. This redundancy was confirmed through our application of the greedy algorithm to a combined set of canonical, data driven, and hypothesis covariance matrices (Figures 3 and 4) Our primary conclusion from this analysis is in line with what the reviewer posited, that there is extensive redundancy among canonical, data driven, and hypothesis covariance matrices that is leading to biologically informative patterns of covariance between site effect sizes to be arbitrarily distributed among them. We additionally found evidence that while the correlations between data driven and environmentally informed (hypothesis) covariance matrices, there is extensive similarity in the matrices that are included in the maximum likelihood models across both phenotypes and in the analysis of the Gulf subpopulation, Midwest subpopulations, and the combined cohort.

``` {#fig-greedy .r}
Greedy algorithm for mash model selection. 
____________________________________________________________________________

@@

matrices.remove(most_likely_matrix) #<4>
most_likely_matrices.append(most_likely_matrix)
maximum_likelihoods.append(likelihood)

lrt_pvalue = 0
iteration = 2
while lrt_pvalue < 0.05: #<5>
    likelihood = -inf
  most_likely_matrix = ‘’
    for matrix in matrices:
        model_matrices = most_likely_matrices + matrix 
        model_likelihood = mash(effects, std.errs, model_matrices)[‘likelihood’]
    if matrix_likelihood > likelihood:
        most_likely_matrix = matrix
        likelihood = matrix_likelihood

  matrices.remove(most_likely_matrix)
  most_likely_matrices.append(most_likely_matrix)
  maximum_likelihoods.append(likelihood)

  lrt_pvalue = likelihood_ratio_test(likelihood, maximum_likelihoods[iteration-1], df = 1)
____________________________________________________________________________
```

1.  Initialize covariance matrices of interest
2.  Initialize lists to store the most likely additional matrix and corresponding maximum likelihood
3.  Obtain model likelihood for a model fit with each matrix
4.  Remove the matrix with the greatest likelihood from the array and store it
5.  Repeat the process of fitting mash models, finding the most likely, pair, trio, etc. (using the most likely matrices from the preceding iteration) until the likelihood ratio test is no longer significant

> ::: {.callout-warning appearance="minimal"}
> \- There appears to be an issue with the construction of the hypothesis-derived covariance matrices. I believe the goal here is to estimate the genetic covariance among gardens based on the phenotypic covariance (P) and the estimated residual covariance (E), where P = G + E. The described approach involves starting with the the phenotypic correlation matrix and then replacing the diagonal with the coefficient of variation. If the diagonal had been replaced with the h2 in each garden, that’d be a valid estimate of G. But as described, it’s likely that you’ll end up with an invalid covariance matrix, one that’s not positive-semi-definite. Instead the appropriate operation would be to both pre- and post-multiply the covariance matrix by a diagonal matrix with the square-root of the CV’s. This maintains a valid covariance-like structure. Or just mean-standardize the traits first and then calculate the covariance of this standardized matrix.
> :::

We now use the narrow-sense heritability as the diagonal for each GxWeather matrix, and clarify this in the Supplementary Methods (SI Appendix, Section S1, first and eighth paragraph - first and last paragraph on page two).

> ::: {.callout-warning appearance="minimal"}
> \- Antagonistic Pleiotropy: Maybe I’m missing the idea of this analysis, but it doesn’t seem to me that this strategy for partitioning loci between antagonistic pleiotropy and differential sensitivity would have equal power to detect each class. It seems that antagonistic pleiotropy should be a subset of differential sensitivity because the latter is defined as simply a change in magnitude. But even if it’s restricted to “different magnitude but same sign”, the practical definition here seems to be that antagonistic pleiotropy is detected when the signs are different and both lfsrs \< threshold (0.05?), while differential sensitivity additionally requires a threshold on the difference in effect sizes (0.4) with no justification for why this size was chosen. This additional criterion will necessarily make the rate of detection different. Even without this, with differential sensitivity one effect size must be much larger than the other in absolute value (which isn’t the case with antagonistic pleiotropy), so differential sensitivity loci will be more likely to pass the lfsr loci than antagonistic pleiotropy loci. Note that previous literature here tried to separate antagonistic pleiotropy from conditional neutrality (ie where in one location the effect was zero), while here the contrast is with differential sensitivity where effects are non-zero in all locations. I’m not aware of previous discussions in the literature of trying to differentiate these two specific classes of GxE variants and not really sure theoretically what the importance would be.
> :::

We believe that the use of the lfsr to detect effects with rank-changing GxE is a key advance in our manuscript, and so we have made changes to the Introduction, Results, and Materials & Methods (lines 957-997; lines 1055-1056), as well as the SI Appendix, to explain and motivate this statistical change.

Our expectation is that in nature, effects will always differ. Statisically, this difference will not always be detectable. Thus, rather than ask "Are these two effects different?" - as we reasonably expect two effects to be, even if this difference cannot be measured - the local false sign rate answers a more meaningful question: Can we be confident in the sign of this effect?

We think that if evolutionary geneticists are interested in antagonistic pleiotropy and rank-changing GxE, then they should use the lfsr to measure their confidence in the sign of the effect, rather than the lfdr to measure their confidence that the effect is different than zero. In the lfsr case, comparisons between antagonistic pleiotropy and conditional neutrality are no longer sensible, as we are no longer doing a statistical test that can robustly detect conditional neutrality (just as the lfdr does not robustly detect antagonistic pleiotropy).

It's true that the threshold for differential sensitivity is arbitrary. We have reworded the manuscript to stress that there is equal power to detect effects of different sign as there are effects of the same sign (lines 994-997; lines 1055-1056). Another class of effects we have in our analysis, that differ from classic antagonistic pleiotropy & conditional neutrality comparisons are effects which are not distinguishable by sign or magnitude. It is not accurate that differentially sensitive loci will be more likely to pass the lfsr, because the loci are first selected by significant lfsr, then divided further into effects that can & cannot be distinguished by magnitude. Thus effects that are significant, but have small magnitude in both conditions will be labeled as not being distinguishable - essentially, these effects do not have detectable GxE. We propose that, if our intent is to detect sign-changing GxE, then we should use an unbiased statistical test to look at loci with and without sign-changing GxE, the lfsr.
